---
title: "R Notebook"
output:
  html_notebook: default
  word_document: default
---

This script downloads and wrangles data from the CDC/Our World in Data to update (i) the vaccination tool, and (ii) the graphs for the vaccine stockpile report.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 
```{r echo=FALSE}
getwd()
# Alex, it looks like it goes right to the vaccinations folder, so if I exclude prefixes, it should work on both machines
```

```{r echo=FALSE}
### Load data from Our World in Data
library(dplyr)
library(zoo)
library(ggplot2)
library(openxlsx)

theme_set( theme_bw() + theme( legend.position="bottom" ) +
             theme( legend.title=element_blank() ) +
             theme(plot.title = element_text(hjust = 0.5) ) +
             theme( text = element_text(size=12) ) )

# Assumptions
# rate of doses lost
# Kaitlin Volk's email on 3/15/2021: NH 0.69%, 0.08% in MA, 0.06% in RI, 0.2% in VT, ME has refused to release statistics, CT="10 doses"
waste_rate <- 0.007 
# average time between 1st and 2nd dose
t1st2nd <- (21+28)/2

# Constants
dfmts <- c('%Y-%m-%d', '%m/%d/%Y')
NEstates <- c('Connecticut','Maine','Massachusetts','New Hampshire','Rhode Island','Vermont')
nper <- 14 # days to smooth supply

# truth URL from the CDC ensemble 
vurl <- 'https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/us_state_vaccinations.csv'
df <- read.csv( vurl )
save(df, file = "data/owid.rda")
```

Calculate daily history
```{r echo=FALSE}
# (current capacity to administer doses per day is in daily_vaccinations - 7-day moving avg)
dcap <- df %>% filter( location %in% NEstates ) %>% 
  group_by( location ) %>%
  mutate( date = as.Date(date, tryFormats = dfmts)
          , total_vaccinations = na.approx( total_vaccinations, na.rm=FALSE )
          , total_distributed = na.approx( total_distributed, na.rm=FALSE )
          , people_vaccinated_per_hundred = na.approx( people_vaccinated_per_hundred, na.rm=FALSE )
          # , daily_available = total_distributed*(1-waste_rate) - total_vaccinations # total availability level IGNORING WASTE
          # , daily_available_avg = rollmean(daily_available, k=nper, fill=NA, align="right") # total availability level IGNORING WASTE
  ) %>% ungroup()
latest_history_date <- max(dcap$date)
print( paste('Latest data from', latest_history_date) )
```
```{r echo=FALSE}
### Totals ###
tmp <- lapply( NEstates, function( state ){
  idx <- dcap$location==state & dcap$date==latest_history_date
  print( paste('Total Administered', state, dcap$total_vaccinations[idx]) )
  print( paste('Total Delivered', state, dcap$total_distributed[idx]) )
  print( paste('People with 1+ Doses', state, dcap$people_vaccinated[idx]) )
  print( paste('People with 2+ Doses', state, dcap$people_fully_vaccinated[idx]) )
  print( paste('Total Administered per 100k', state, dcap$total_vaccinations_per_hundred[idx]*1000) )
  print( paste('Total Delivered per 100k', state, dcap$distributed_per_hundred[idx]*1000) )
  print( paste('People with 1+ Doses per 100k', state, dcap$people_vaccinated_per_hundred[idx]*1000) )
  print( paste('People with 2+ Doses per 100k', state, dcap$people_fully_vaccinated_per_hundred[idx]*1000) )
} )
```
CDC population numbers
```{r echo=FALSE}
pop <- dcap %>% mutate( pop1 = total_vaccinations/(total_vaccinations_per_hundred/100),
                        pop2 = total_distributed/(distributed_per_hundred/100) ) %>%
      group_by( location ) %>%
      summarise( m=mean(pop1, na.rm=TRUE), s1=sd(pop1, na.rm=TRUE)
                , m2=mean(pop2, na.rm=TRUE), s2=sd(pop2, na.rm=TRUE), .groups='drop_last' )
print(pop)
```
Plot daily_vaccinations (it is already 7-day smoothed in the original data)
```{r echo=FALSE}
dp <- dcap %>% filter( date >= "2021-03-01" ) %>% select(date,location,daily_vaccinations) 
ggplot( dp, aes(x = date, y=daily_vaccinations) ) +
  facet_wrap(~ location, nrow=2, scale='free_y' ) +
  geom_line() + 
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = '%b%e' ) +
  ylab('Daily Vaccinations') +
#  ggtitle( paste('Latest data from', latest_history_date) ) +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")
                     , expand = c(0, 0), limits = c(0, NA))
```
Plot distribution in the last weeks
```{r echo=FALSE}
dp <- dcap %>% filter( date >= "2020-01-01" ) %>% # select(date,location,daily_vaccinations) 
                group_by( location ) %>%
                mutate( dd_dist  = total_distributed - lag(total_distributed, 7) )
ggplot( dp, aes(x = date, y=dd_dist) ) +
  facet_wrap('~ location', nrow=2, scale='free_y' ) +
  geom_line() + 
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = '%b%e' ) +
  ylab('Doses distributed per week') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "") #, expand = c(0, 0), limits = c(0, NA)
                     )
```
Forecasting: 
```{r echo=FALSE}
# Forecast function
fcastfn <- function( l, c, al, be ){
  # bt = NULL: function tries to find
  # bt = FALSE: exponential smoothing only
  
  fdays <- as.Date('2021-10-15') - latest_history_date # number of days to forecast
  
  dt <- dcap %>% filter( location == l ) %>% 
    select( all_of(c('date',c)) ) 
  z <- read.zoo( dt )
  zz <- z
  time(zz) <- seq_along(time(zz))
  # forecast model
  sm <- HoltWinters( as.ts(zz[-(1:(nper-1))]), alpha=al, beta=be, gamma=FALSE )
#  sm <- HoltWinters( as.ts(zz[-(1:(nper-1))]), gamma=FALSE, optim.start=c(alpha=0.1,beta=0.1) ) 
  # tmp <- HoltWinters( as.ts(zz), alpha=0.5, beta=FALSE, gamma=FALSE )
  print( paste('location', l, 'alpha', sm$alpha, 'beta', sm$beta, 'level', sm$coefficients['a'], 'trend', sm$coefficients['b']) )
  pr <- predict(sm, fdays, prediction.interval=TRUE, level = 0.95)
#  plot(sm, pr)
  dates <- seq(max(dt$date)+1, by = "day", length.out = fdays)
  prdf <- data.frame( dates, pr ) 
  prdf[prdf < 0] <- 0    # override negative predictions
  cols <- paste0(c, '_', colnames(prdf[1,-1]))
  colnames(prdf) <- c('date',cols)
  
#  prdf <- prdf %>% tidyr::pivot_longer( cols = -date, names_to="s", values_to = "value",  values_drop_na = TRUE )
  
  # bind history and prediction
  # res <- dt %>% rename( 'value' = c ) %>%
  #   mutate( 's' = c ) %>%
  #   bind_rows( prdf ) %>%
  #   mutate( 'location' = l )
  res <- prdf %>% mutate( 'location' = l )
}
# Forecast daily supply level of doses
#tmp <- lapply( NEstates, fcastfn, 'daily_available_avg' )
print( paste('Latest data from', latest_history_date) )
```
Forecasts
```{r echo=FALSE}
f <- list( )
beta_vaccinations <- 0.03 # Beta forced because of recent spikes in NH
```
Forecast Cumulative number of doses **distributed**
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
tmp <- lapply( NEstates, fcastfn, 'total_distributed', NULL, NULL ) 
f[[1]] <- bind_rows( tmp )
```
Forecast Cumulative number of doses **administered**
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
tmp <- lapply( NEstates, fcastfn, 'total_vaccinations', NULL, beta_vaccinations ) 
f[[2]] <- bind_rows( tmp ) %>% select(-date,-location)
```
Forecast **Daily** number of doses **administered**
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
# Forecast daily vaccinations
tmp <- lapply( NEstates, fcastfn, 'daily_vaccinations', NULL, beta_vaccinations ) 
f[[3]] <- bind_rows( tmp ) %>% select(-date,-location)
```
Forecast people vaccinated per capita
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
tmp <- lapply( NEstates, fcastfn, 'people_vaccinated_per_hundred', NULL, beta_vaccinations )
f[[4]] <- bind_rows( tmp ) %>% select(-date,-location)
```
Bind history and prediction
```{r echo=FALSE}
fcast <- bind_cols( f )
latest_fcast_date <- max(fcast$date)

#dcap_s <- dcap %>% select( date, location, daily_vaccinations,   daily_available )
dcap_s <- dcap %>% select( date, location, daily_vaccinations,   total_distributed,    total_vaccinations,  people_vaccinated_per_hundred )
fcast_s <- fcast %>% select(date,location,daily_vaccinations_fit,total_distributed_fit,total_vaccinations_fit,people_vaccinated_per_hundred_fit)
colnames(fcast_s) <- colnames(dcap_s)
dfw <- bind_rows( dcap_s, fcast_s ) %>%
  mutate( daily_available = total_distributed*(1-waste_rate) - total_vaccinations ) # total availability level
```
Plot Forecasted Cumulative number of doses distributed & administered
```{r echo=FALSE}
print( paste0('\'Availability\' = [Cumulative number of doses distributed] - [Cumulative number of doses administered] - [waste(', waste_rate*100, '% of total)]') )
```
```{r echo=FALSE}
colors <- c('total_distributed','total_vaccinations','daily_available')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>%
  tidyr::pivot_longer( cols = all_of(colors), names_to="s", values_to="value", values_drop_na=TRUE )
cnames <- c('Distributed','Administered','Availability')
dp <- dp %>% mutate( n = NA )
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)]
ggplot( dp, aes(x = date, y=value, color=n) ) +
  facet_wrap('~ location', nrow=2, scale='free_y' ) +
  geom_line() + 
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  ggtitle( paste('Latest data from', latest_history_date) ) +
  theme(axis.title.x = element_blank()) +
  scale_x_date( date_labels = '%b%e' ) +
  ylab('Cummulative Number of Doses') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")
                     , expand = c(0, 0), limits = c(0, NA))
```
Plot Forecasted Cumulative number of people vaccinated per hundred
```{r echo=FALSE}
ggplot( dp, aes(x = date, y=people_vaccinated_per_hundred) ) +
  facet_wrap('~ location', nrow=2 ) +
  geom_line() + 
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  ggtitle( paste('Latest data from', latest_history_date) ) +
  theme(axis.title.x = element_blank()) +
  scale_x_date(date_labels = '%b%e' ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, NA))
```
## MAXIMUM CAPACITY
```{r echo=FALSE}
# calculate maximum capacity to administer doses per day (assumed constant)
fname <- 'data/vaccine capacity.xlsx'
#fname <- '/Users/aligo/Downloads/FEMA recovery data/vaccine capacity.xlsx'
sites <- read.xlsx( fname, sheet = 1, detectDates=TRUE )
sitecap <- read.xlsx( fname, sheet = 2, detectDates=TRUE )

sc <- left_join(sites, sitecap, by=c('date','location','type')) %>%
  group_by(date, location) %>% summarise( cap = sum( Ni * Ci ), .groups='drop_last' )

#dfw <- dfw %>% mutate( cap = NA )
for ( state in NEstates )
  dfw$cap[dfw$location==state #& dfw$date>latest_history_date-5
          ] <- sc$cap[sc$location==state]

colors <- c('daily_available','daily_vaccinations','cap')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>%
#  mutate( supply_daily = zoo::rollmean(supply_daily, k=7, fill=0, align="right") ) %>% 
  tidyr::pivot_longer( cols = all_of(colors)
                       , names_to="s" 
                       , values_to = "value",  values_drop_na = TRUE )
cnames <- c('Supply level','Doses administered','Max vaccination capacity (FAKE)')
dp <- dp %>% mutate( n = NA )
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)]
ggplot( dp, aes(x = date, y=value, color=n) ) +
  facet_wrap('~ location', nrow=2, scale='free_y' ) +
  geom_line() +
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  theme(axis.title.x = element_blank()) +
  ggtitle( paste('Latest data from', latest_history_date) ) +
  ylab('Doses per day') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = ""))
```
## calculate derivative of daily_vaccinations and totals per 100K people
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date) )
# derivative of daily_vaccinations
dfw <- dfw %>% group_by( location ) %>%
  mutate( dd_vaccinations = (daily_vaccinations-lag(daily_vaccinations))/as.numeric(date-lag(date)) 
          , daily_vaccinations_per_million=NA
          , total_vaccinations_per_hundred=NA
          , distributed_per_hundred=NA )

# totals per 100K people (history and forecast)
for ( state in NEstates ){
  popst <- pop$m[pop$location==state]
  idx <- dfw$location==state
  dfw$daily_vaccinations_per_million[idx] <- dfw$daily_vaccinations[idx] / popst * 1e6
  dfw$total_vaccinations_per_hundred[idx] <- dfw$total_vaccinations[idx] / popst * 100
  dfw$distributed_per_hundred[idx] <- dfw$total_distributed[idx] / popst * 100
}
```

```{r echo=FALSE}
## Forecast total availability and vax per state (for bar graphs) ##
comp <- function( d ){
  # This function calculates availability and required doses for the next d days
  
  # current stockpile: daily_available = total_distributed*(1-waste_rate) - total_vaccinations, all at latest_history_date
  # Number of people who current need a 2nd dose: demand2nd = people_vaccinated - people_fully_vaccinated, all at latest_history_date
  # assuming everyone in demand2nd will get the 2nd dose within 24 days (average between Moderna and Pfizer), then 7/24 of them will get the shot within one week
  stock <- dcap %>% filter( date == latest_history_date ) %>%
    mutate( demand2nd = ( people_vaccinated - people_fully_vaccinated ) * d/t1st2nd ,
            daily_available = total_distributed*(1-waste_rate) - total_vaccinations ) %>% # total availability
    select( location, daily_available, demand2nd ) 

  # Number of people who will get 1st dose next week: demand1st1wk = sum[daily_vaccinations(next d days)] - demand2nd
  dvac <- dfw %>% filter( between( date, latest_history_date+1, latest_history_date+d ) ) %>%
    group_by( location ) %>%
    summarise( dvac = sum(daily_vaccinations), .groups='drop_last' )
  stock <- left_join(stock, dvac, by='location') %>%
    mutate( demand1st = dvac - demand2nd )
  stock[stock < 0] <- 0
  
  # Number of doses that will be distributed next week = total_distributed(1wk ahead) - total_distributed(current)
  # = daily_available(1wk ahead) + total_vaccinations(1wk ahead) - daily_available(current) - total_vaccinations(current)
  # = daily_available(1wk ahead) - daily_available(current) + sum[daily_vaccinations(next 7 days)]
  # daily_available(1wk ahead)
  dav <- dfw %>% filter( date==latest_history_date+d ) %>% select(location,daily_available) %>% rename( dav=daily_available ) 
  stock <- left_join(stock, dav, by='location') %>%
    mutate( distrib = dav - daily_available + dvac )

  dp <- stock %>% select( -dvac,-dav ) %>%
    tidyr::pivot_longer( cols = -location, names_to="variable" ) %>%
    mutate( cat = 'Available')
  
  # which stack each total shows in
  dp[grepl('^demand',dp$variable),'cat'] <- 'Needed'
  
  # code to choose colors to use
  # hex_codes <- scales::dichromat_pal("BluetoOrange.10")(10)   # Identify hex codes
  # dichromat_pal("BluetoOrange.10")(5)
  # dichromat_pal("DarkRedtoBlue.12")(12)
  # scales::show_col(hex_codes) 
  
  # legend (colors)
  # legend must be sorted in the order they should appear in the stack
  bars <- c('addl forecasted\ndistribution','current\navailability','1st dose\nforecast','2nd dose\nforecast')
  leg <- c(bars[2],bars[1],bars[3],bars[4])
  colors<- c('#3399FF','#66CCFF', '#FF9933', '#FF5500')
  dp[dp$variable=='daily_available','variable'] <- leg[1]
  dp[dp$variable=='distrib','variable'] <- leg[2]
  dp[dp$variable=='demand1st','variable'] <- leg[3]
  dp[dp$variable=='demand2nd','variable'] <- leg[4]
  
#  print( dp )
  p <- ggplot(dp, aes(x=cat, y=value, fill=factor(variable, levels=bars))) + 
    geom_bar(stat = 'identity', position = 'stack') + 
    facet_wrap('~ location', nrow=2, scale='free_y' ) +
    #  scale_fill_discrete(breaks=leg) +
    scale_fill_manual(values=colors, breaks=leg) +
    theme(axis.title.x = element_blank()) +
    ylab('Number of doses') +
    scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")) + 
    ggtitle( paste0(d, '-Day Projection of Doses Available and Needed')) + theme(plot.title = element_text(hjust = 0.5) )
  
  return( list(stock, p) )
}
```

# WEEKLY REPORT STARTS HERE

# FEMA-R1 Vaccine Assessment
```{r echo=FALSE}
print( paste('Latest data from', latest_history_date))
```
## 7-day forecast of dose availability and administration
```{r echo=FALSE}
tmp <- comp( 7 )
stock7d <- tmp[[1]]
print( stock7d )
```
```{r echo=FALSE}
tmp[[2]]
```
Figure 1. 7-day forecast of availability and administration of doses for the New England states. Source: CDC/Our World in data

# CALCULATIONS for Shiny App
```{r}
# capacity assumptions
sc <- left_join(sites, sitecap, by=c('date','location','type')) 

# main dataframes
colors <- c('daily_vaccinations','cap')
cnames <- c('Doses administered','Max vaccination capacity')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>% 
    tidyr::pivot_longer( cols = all_of(colors), names_to="s" 
                         , values_to = "value",  values_drop_na = TRUE )
dp <- dp %>% mutate( n = NA )
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)]

# per capita values per state and Region 1 average
dpp <- data.frame()
for( var in c('total_vaccinations_per_hundred','people_vaccinated_per_hundred') ){
  tmp <- dfw %>% select( c('date','location',all_of(var)) ) %>%
                       tidyr::pivot_wider( names_from='location', values_from=var ) %>%
                       mutate( avgr1 = (.[[2]]+.[[3]]+.[[4]]+ .[[5]]+ .[[6]]+ .[[7]]) /6 ) %>%
                       tidyr::pivot_longer( cols=all_of(NEstates), names_to="location" 
                                            , values_to = "value",  values_drop_na=FALSE ) %>%
                       tidyr::pivot_longer( cols=c(value,avgr1), names_to="type" 
                                            , values_to=var, values_drop_na=FALSE ) 
  if ( nrow(dpp) == 0 )
    dpp <- tmp
  else
    dpp[,var] <- tmp[,var]
}
dpc <- dpp %>% filter( date <= (latest_history_date+14) )

# Perc people vaccinated over 16 yr old
perc_pop_over16 <- data.frame( state = NEstates,
                               perc_over16 = c(.821, .837, .827, .836, .83, .84) )
for ( st in NEstates ){
  idx <- dpp$location == st
  dpp$people_vaccinated_per_hundredover16[idx] <- dpp$people_vaccinated_per_hundred[idx] / perc_pop_over16$perc_over16[perc_pop_over16$state==st]
}
  
# dates where given vars will exceed given percentages of population
calc_threshold_dates <- function(st, vartype){
  tmp <- lapply( c('people_vaccinated_per_hundred','people_vaccinated_per_hundredover16'), function(colref){
    thrs <- c(70, 90)
    dates <- sapply( thrs, function(threshold){
      datetmp <- dpp %>% filter( type==vartype & location==st & !!as.symbol(colref)>=threshold )
      stopifnot( nrow(datetmp) > 0 )
      return( min( datetmp$date ) )
    } )
    return( data.frame( location=st, type=vartype, colref, threshold=thrs, date=as.Date(dates) ) )
  } ) %>% bind_rows()
}
datesdf <- lapply( NEstates, calc_threshold_dates, 'value' ) %>% bind_rows() %>% tidyr::pivot_wider( names_from=colref, values_from=date )
datesdfavg <- calc_threshold_dates( 'Connecticut', 'avgr1' )

# forecasted growth in daily vaccinations
dvacgr <- dfw %>% group_by( location ) %>%
                summarise( b = last(dd_vaccinations), .groups='drop_last' )
```
```{r}
# test only
tmp <- dpp %>% filter( type=='value' & location==NEstates[4]  )
tmp
```


## SAVE data for Shiny App app.R and for in-transit graph of pre_proc_Tiberius.Rmd
```{r echo=FALSE}
print( paste('SAVING Latest data from', latest_history_date) )
#save(dfw, dp, dpp, dpc, datesdf, datesdfavg, dvacgr, latest_history_date, sc, pop, NEstates, stock7d,  file='/Users/aligo/git-repos/FEMA/vaccinations/app.RData')
save(dfw, dp, dpp, dpc, datesdf, datesdfavg, dvacgr, latest_history_date, sc, pop, NEstates, stock7d,  file='app/app.RData')
```

## 14-day forecast of availability and administration of doses
```{r echo=FALSE}
tmp <- comp( 14 )
print( tmp[[1]] )
```
```{r echo=FALSE}
tmp[[2]]
```
Figure 2. 14-day forecast of availability and administration of doses for the New England states. Source: CDC/Our World in data.

## Forecasting method
```{r echo=FALSE}
colors <- c('total_distributed','total_vaccinations','daily_available')
dp <- dfw %>% filter( date <= (latest_history_date+14) ) %>%
  tidyr::pivot_longer( cols = all_of(colors), names_to="s", values_to="value", values_drop_na=TRUE )
cnames <- c('Cumulative Distributed','Cumulative Administered','Availability')
dp <- dp %>% mutate( n = NA )
dp$n[dp$s %in% colors] <- cnames[match(dp$s,colors)]
ggplot( dp, aes(x = date, y=value, color=n) ) +
  facet_wrap('~ location', nrow=2, scale='free_y' ) +
  geom_line() + 
  geom_vline(xintercept=latest_history_date, linetype="dashed", 
             color = "gray") +  
  theme(axis.title.x = element_blank(), text = element_text(size=11) ) +
  scale_x_date(date_labels = '%b%e' ) +
  ylab('Number of Doses') +
  scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")
                     , expand = c(0, 0), limits = c(0, NA))
```
Figure 3. Forecast of cumulative number of distributed doses, cumulative number of doses administered, and availability of doses for the New England states. Source: CDC/Our World in data, Feb 21.
