---
title: "R Notebook"
output:
  html_notebook: default
  word_document: default
---
This is to prepare data for the weekly report/forecast on vaccine availability (stockpile) and vaccinations

```{r echo=FALSE}
### Load data from Our World in Data
library(dplyr)
library(zoo)
library(ggplot2)

theme_set( theme_bw() + theme( legend.position="bottom" ) + theme( legend.title=element_blank() ) + theme(plot.title = element_text(hjust = 0.5) ) + theme( text = element_text(size=12) ) )
```

Wrangling of Tiberius data of orders and shipments
```{r}
load("app/app.RData")
fname <- '/Users/aligo/Downloads/FEMA recovery data/Data_tiberius/csv/Orders+Shipments+and+Deliveries+-+Reporting+Dataset.csv'
orders <- read.csv( fname )
```

Total shipments per NDC
```{r}
ndc <- orders[,c('ndc','ndc_description')]
print( 'list of NDC descriptions:' ) # see meaning in http://publichealth.lacounty.gov/acd/docs/COVID-19VaccineProductInfoGuide.pdf
```
```{r}
unique( ndc ) 
```

Total shipments per awardee
```{r}
vxord <- orders %>% filter( ndc %in% c('59267-1000-02','80777-0273-99','59676-0580-15') & 
                              shipping_acl_awardee %in% c('Connecticut','Maine','Massachusetts','New Hampshire','Rhode Island','Vermont') )
acat <- unique( vxord$awardee_category )
print( 'list of awardee categories:' )
print(acat)
```
```{r echo=FALSE}
dp <- vxord %>% group_by(awardee_category,awardee_id,awardee) %>% 
  summarise( n = n(), .groups="drop_last" ) %>% ungroup()
# Bar graph
ggplot(dp, aes(x=awardee, y=n, fill=awardee_category) ) +
    geom_bar(stat="identity", position=position_dodge()) +
    coord_flip()
```
Total shipments per shipping awardee
```{r echo=FALSE}
acat <- unique( vxord$shipping_awardee_category )
print( 'list of shipping awardee categories:' )
print(acat)
```
```{r echo=FALSE}
dp <- vxord %>% group_by(shipping_acl_awardee_id,shipping_acl_awardee) %>% summarise( n = n(), .groups="drop_last" )
# Bar graph
ggplot(dp, aes(x=shipping_acl_awardee, y=n) ) +
    geom_bar(stat="identity", position=position_dodge()) +
    theme(legend.position="none") +
    coord_flip()
```
Distribution of delay from order to delivery
```{r}
vxord <- vxord %>% mutate( first_delivery_date = as.Date(substr(first_delivery_timestamp,1,10), format="%Y-%m-%d") )
dp <- vxord %>% mutate( delay = first_delivery_date - as.Date(order_approved_date, format="%Y-%m-%d") )
ggplot(dp, aes(x=delay)) + geom_histogram(binwidth=1)
```
Get cummulative doses in transit and distributed from Tiberius
```{r}
#latest_history_date <- as.Date( '2021-03-22' )
datelst <- seq(as.Date("2020-12-10"), to=latest_history_date, by = "day")
tmp <- lapply( datelst, function(d){
  dord <- vxord %>%
    group_by( shipping_acl_awardee, vax_manufacturer, ndc ) %>%
    summarise( in_transit_total = sum( doses_approved[order_approved_date<=d & (is.na(first_delivery_date) | first_delivery_date>d)] )
               , total_distributed = sum( doses_approved[first_delivery_date<=d], na.rm=TRUE )
               , .groups='drop_last' ) %>% mutate( date = d ) %>%
    select( shipping_acl_awardee, vax_manufacturer, ndc, date, in_transit_total, total_distributed )
}) 
dord <- bind_rows( tmp ) %>% ungroup()
```
Total distributed in latest_history_date - sanity check with CDC data
```{r}
dord %>% filter( date == latest_history_date ) %>%
    group_by( shipping_acl_awardee ) %>%
    summarise( total_distributed = sum( total_distributed ), .groups='drop_last' ) 
```
Plot history 
```{r}
for( vx in unique(dord$vax_manufacturer) ){
  dp <- dord %>% filter( vax_manufacturer == vx ) %>%
                tidyr::pivot_longer( cols=c(in_transit_total, total_distributed), names_to="s", values_to = "value",  values_drop_na = TRUE )
  p <- ggplot( dp, aes(x=date, y=value, color=s) ) +
            facet_wrap(~ shipping_acl_awardee, ncol=3, scales='free_y' ) +
            geom_line( size=1.5 ) +
            geom_vline(xintercept=latest_history_date, linetype="dashed", color="gray", size=1) +
            theme(axis.title.x = element_blank()) +
            ggtitle( vx ) +
            scale_x_date(date_labels = '%b%e' ) +
            ylab( 'Total doses distributed and in transit' ) +
            scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "") )
  print( p )
}
```
# Doses delivered per week per State and per Manufacturer
```{r}
vxordd <- vxord %>% 
          group_by( shipping_acl_awardee, vax_manufacturer, first_delivery_date ) %>%
          summarise( day_deliv = sum(doses_delivered), .groups='drop_last' ) %>% 
          group_by( shipping_acl_awardee, vax_manufacturer ) %>%
          mutate( day_deliv_avg = zoo::rollmean(day_deliv, k=7, na.pad=TRUE, align="right") ) 
tmp <- vxordd %>% filter( shipping_acl_awardee=='Massachusetts', substr(vax_manufacturer,1,6) == "Pfizer")

ggplot( vxordd, aes(x=first_delivery_date, y=day_deliv, color=vax_manufacturer) ) +
            facet_wrap(~ shipping_acl_awardee, ncol=3, scales='free_y' ) +
            geom_point( size=0.5 ) +
            geom_smooth(method = lm, formula = y ~ splines::ns(x,2), se = FALSE) +
            scale_x_date( name=NULL, date_labels = '%b%e' ) +
            ylab( 'Doses delivered per day' ) +
            scale_y_continuous(limits=c(0,NA), label=scales::unit_format(unit="K", scale=1e-3, sep="") )

ggplot( vxordd, aes(x=first_delivery_date, y=day_deliv_avg, color=vax_manufacturer) ) +
            facet_wrap(~ shipping_acl_awardee, ncol=3, scales='free_y' ) +
            geom_point() +
            scale_x_date( name=NULL, limits=c(as.Date('2021-03-15'),NA), date_labels = '%b%e' ) +
            ylab( 'Doses delivered per day (7-day moving average)' ) +
            scale_y_continuous(limits=c(0,NA), label=scales::unit_format(unit="K", scale=1e-3, sep="") )
```

### Stacked + percent bars
```{r, fig.width = 8, fig.height = 4}
dlabels = seq(as.Date("2021-01-25"), as.Date("2021-04-26"), "2 weeks")
vxordw <- vxord %>% filter( first_delivery_date>"2020-01-01" )
vxordw$ft <- sapply( vxordw$first_delivery_date, function(x) dlabels[which(abs(as.numeric(dlabels-x)) == min(abs(as.numeric(dlabels-x))))[1]] )
vxordw$ft <- as.Date(vxordw$ft)

vxordw <- vxordw %>% 
#          mutate( day = as.numeric(format(first_delivery_date, format='%d')), ft = paste0( format(first_delivery_date, format='%m'), '(', format(first_delivery_date, format='%b'), ') ', ifelse(day>15,'H2','H1') ) ) %>% # fortnite
#          group_by( ft ) %>% mutate( s = min(first_delivery_date) ) %>%
          group_by( ft, shipping_acl_awardee ) %>% mutate( dtot = sum(doses_delivered) ) %>% 
          group_by( ft, shipping_acl_awardee, dtot, vax_manufacturer, ndc ) %>%
          summarise( ft_deliv = sum(doses_delivered), .groups='drop_last' ) %>% 
          mutate( ft_deliv_ratio = ft_deliv / dtot,
                  dtotlabel = paste0(round(dtot/1000),'K') )

ggplot(vxordw, aes(fill=vax_manufacturer, y=ft_deliv_ratio, x=ft)) + 
    facet_wrap(~ shipping_acl_awardee, ncol=3 ) +
    geom_col() +
    geom_text( aes( y = 1.06, label = dtotlabel ), size=2.5 ) +
    scale_x_date( name=NULL, limits=c(as.Date('2021-02-26'),as.Date('2021-04-29')), breaks=dlabels, date_labels = '%b%e' ) + 
    scale_y_continuous( name='Doses delivered per 14-day period', limits=c(0,NA), label=scales::percent )
```


# Load data about doses administered from Tiberius
```{r}
fname <- '/Users/aligo/Downloads/FEMA recovery data/Data_tiberius/csv/County+Vaccine+Administration+-+Reporting+Dataset.csv'
df <- read.csv( fname )
```
```{r}
unique(df$awardee)
# this shows that awardee is not good to get state totals
# tmp <- df %>% group_by( awardee ) %>% summarise( t = max(cumulative_total_doses_administered) )
# print.data.frame(tmp)
dadm <- df %>% mutate( date=as.Date(date, tryFormats=c('%Y-%m-%d')), state=awardee, stfips=substr(county_fips,1,2) ) 
fedstuff <- c("Veterans Health","Dept of Defense","Indian Health Svc","Bureau of Prisons","Dept of State")
for (fed in fedstuff){
  idx <- dadm$awardee==fed
  dadm$state[idx] <- dadm$stfips[idx]
}
NEstates <- c("Connecticut","Maine","Massachusetts","New Hampshire","Rhode Island","Vermont")
NEfips <- c('09','23','25','33','44','50')
for (i in 1:6)
  dadm$state[dadm$state==NEfips[i]] <- NEstates[i]

dadm <- dadm %>% filter( state %in% NEstates ) 
```

Check total vaccinations from Tiberius
```{r}
dadms <- dadm %>% 
  group_by( state, vaccine, ndc, date ) %>%
  summarise( total_vaccinations = sum( cumulative_total_doses_administered, na.rm=FALSE )
             , total_1st = sum( cumulative_first_doses_administered, na.rm=FALSE )
             , total_2nd = sum( cumulative_second_doses_administered, na.rm=FALSE )
             , total_unc = sum( cumulative_dose_unclassified, na.rm=FALSE )
    , .groups='drop_last' ) %>% ungroup()

dadms[dadms$date==latest_history_date,]
```
Merge cummulative distribution and vaccinations from Tiberius
```{r}
dcap <- left_join( dord, dadms, by=c('shipping_acl_awardee'='state', 'ndc'='ndc', 'date'='date') ) %>%
  rename( state = shipping_acl_awardee )
```

Forecasting: 
```{r echo=FALSE}
# Forecast function
fcastfn <- function( st, col, vxndc, al=NULL, be=NULL ){
  # st: state name
  # d: number of latest days to ignore for forecast
  # col: variable to forecast
  # vx: vaccine manufacturer
  # be = NULL: function tries to find
  # be = FALSE: exponential smoothing only
  dt <- dcap %>% filter( state == st & ndc == vxndc # & date < (latest_history_date-d) 
                         ) %>% select( all_of(c('date',col)) ) 
  z <- read.zoo( dt )
  zd <- coredata(z)
  d <- 0
  repeat{
    if ( zd[length(zd)] != zd[length(zd)-d] )
      break
    else
      d <- d+1
  }
  zz <- z[1:(length(z)-d+1)]
  time(zz) <- seq_along(time(zz))
  # forecast model
  sm <- HoltWinters( as.ts(zz), alpha=al, beta=be, gamma=FALSE )
  print( paste(st,-d, 'alpha',sm$alpha, 'beta',sm$beta, 'level',sm$coefficients['a'], 'trend',sm$coefficients['b']) )
  pr <- predict(sm, 14, prediction.interval=TRUE, level = 0.95)
  dates <- seq(max(dt$date)+1, by = "day", length.out = 14)
  prdf <- data.frame( dates, pr ) 
  prdf[prdf < 0] <- 0    # override negative predictions
  cols <- paste0(col, '_', colnames(prdf[1,-1]))
  colnames(prdf) <- c('date',cols)
  
  res <- prdf %>% mutate( state = st, ndc=vxndc ) # , var_name = col
}
```

Forecast Cumulative number of doses **distributed** and **administered**
```{r}
print( paste('Latest data from', latest_history_date) )
cols <- c('in_transit_total','total_distributed','total_vaccinations','total_1st','total_2nd','total_unc')
vaccines <- c('59267-1000-02','80777-0273-99') # unique( dcap$ndc ) # c('Moderna','Pfizer')
tmp <- lapply( vaccines, function(vxndc){
  print( paste('vaccine ndc',vxndc) )
  f <- list( )
  for( col in cols ){
      print( col )
      res <- lapply( NEstates, fcastfn, col, vxndc )
      # final days to ignore in forecast 
      #  for CT, ME, MA, NH, RI, VT
#      d <- c( 4,  8,  4,  2,  3,  7 )
#      res <- mapply( fcastfn, NEstates, col, vx, SIMPLIFY=FALSE )
      i <- length(f) + 1
      if ( i == 1 ){
        f[[i]] <- bind_rows( res )
      }
      else
        f[[i]] <- bind_rows( res ) %>% select(-date,-state,-ndc)
  }
  return( bind_cols(f) )
} )
fcast <- bind_rows( tmp )

# Bind history and prediction
dcap_s <- dcap %>% select( state, ndc, date, all_of(cols) )
fcast_s <- fcast %>% select( state, ndc, date, all_of( paste0(cols,'_fit') ) )
colnames(fcast_s) <- colnames(dcap_s)
dfw <- bind_rows( dcap_s, fcast_s ) 
```
Plot history and prediction of cummulative vaccinations per dose and manufacturer
```{r}
for( vxndc in vaccines ){
  dp <- dfw %>% filter( ndc == vxndc ) %>%
                tidyr::pivot_longer( cols = all_of(cols), names_to="s", values_to = "value",  values_drop_na = TRUE )
  p <- ggplot( dp, aes(x=date, y=value, color=s) ) +
            facet_wrap(~ state, ncol=3, scales='free_y' ) +
            geom_line( size=1 ) +
#            scale_color_manual( name=NULL, values=c('red','grey'), breaks=c('value','avgr1'), labels=c('Total in state','Region 1 average') ) +
            geom_vline(xintercept=latest_history_date, linetype="dashed", color="gray", size=1) +
            theme(axis.title.x = element_blank()) +
            ggtitle( vxndc ) +
            scale_x_date(date_labels = '%b%e' ) +
            ylab( 'Total doses administered' ) +
            scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "") )
  print( p )
}
```
# Comparison forecast of Stockpile with forecast of vaccinations
## (bar graphs for weekly report) ##
```{r}
comp <- function( d ){
  # This function calculates availability and required doses for the next d days

  # Assumptions
  waste_rate <- 0.007  # rate of doses lost

  # current stockpile: daily_available = total_distributed*(1-waste_rate) - total_vaccinations, all at latest_history_date
  # Number of people who will need a 2nd dose within d days: demand2nd = total_2nd(latest_history_date+d) - total_2nd(latest_history_date)
  # Similarly with Number of people who will get 1st dose
  # Number of doses that will be distributed within d days = total_distributed(latest_history_date+d) - total_distributed(latest_history_date)
  dcurnt <- dcap %>% filter( date == latest_history_date )
  dfcast <- dfw %>% filter( date == (latest_history_date + d) )
  stock <- left_join( dcurnt, dfcast, by=c('state','ndc'), suffix = c(".c", ".f") ) %>%
            mutate( daily_available = total_distributed.c *(1-waste_rate) - total_vaccinations.c
                    , demand2nd = total_2nd.f - total_2nd.c
                    , demand1st = total_1st.f - total_1st.c
                    , distrib = total_distributed.f - total_distributed.c ) %>% 
            select( state, vax_manufacturer, daily_available, demand2nd, demand1st, distrib ) 
  #stock[stock < 0] <- 0
  
  # total all vaccine manufac
  stockall <- stock %>% group_by( state ) %>%
              summarise( daily_available = sum( daily_available, na.rm=TRUE ), demand2nd = sum( demand2nd, na.rm=TRUE ), demand1st = sum( demand1st, na.rm=TRUE ), distrib = sum( distrib, na.rm=TRUE ), .groups='drop_last' ) %>%
              mutate( vax_manufacturer = 'All' )
  
  dp <- bind_rows( list(stock, stockall) ) %>% 
    tidyr::pivot_longer( cols = c(-state,-vax_manufacturer), names_to="variable" ) %>%
    mutate( cat = 'Available')
  
  # which stack each total shows in
  dp[grepl('^demand',dp$variable),'cat'] <- 'Needed'
  
  # legend (colors)
  # legend must be sorted in the order they should appear in the stack
  bars <- c('addl forecasted\ndistribution','current\navailability','1st dose\nforecast','2nd dose\nforecast')
  leg <- c(bars[2],bars[1],bars[3],bars[4])
  colors<- c('#3399FF','#66CCFF', '#FF9933', '#FF5500')
  dp[dp$variable=='daily_available','variable'] <- leg[1]
  dp[dp$variable=='distrib','variable'] <- leg[2]
  dp[dp$variable=='demand1st','variable'] <- leg[3]
  dp[dp$variable=='demand2nd','variable'] <- leg[4]
  
  plotbars <- function( dp, vx ){
    ggplot(dp, aes(x=cat, y=value, fill=factor(variable, levels=bars))) + 
      geom_bar(stat = 'identity', position = 'stack') + 
      facet_wrap(~ state, nrow=2, scale='free_y' ) +
      scale_fill_manual(values=colors, breaks=leg) +
      theme(axis.title.x = element_blank()) +
      ylab('Number of doses') +
      scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")) + 
      ggtitle( paste0(vx, ': ', d, '-Day Projection of Doses Available and Needed') ) + 
      theme(plot.title = element_text(hjust = 0.5) )
  }
  pvx <- list()
  for( vx in unique(dp$vax_manufacturer) ){
    dpv <- dp %>% filter( vax_manufacturer == vx )
    pvx[[length(pvx)+1]] <- plotbars( dpv, vx )
  }
  return( pvx )
}
```

## 7-day forecast of dose availability and administration
```{r echo=FALSE}
res7d <- comp( 7 )
#stock7d <- tmp[[1]]
#print( stock7d )
```
```{r}
res7d[[2]]
```
```{r}
res7d[[3]]
```
```{r}
res7d[[4]]
```

Calculation of doses in transit on a specific ref_date: sum of doses_approved [ order_approved_date <= latest_history_date & (first_delivery_date == NA | first_delivery_date > latest_history_date) ]
```{r}
print( paste('Total Doses in Transit on', latest_history_date) )
```
```{r}
dp <- vxord %>% group_by( shipping_awardee_category, shipping_acl_awardee, vax_manufacturer ) %>%
  summarise( in_transit_total = sum( doses_approved[order_approved_date<=latest_history_date & (is.na(first_delivery_date) | first_delivery_date>latest_history_date)] ), .groups='drop_last' )

tmp <- stock7d %>% mutate( l = 'addl forecasted distribution' )
tmp
ggplot(dp) +
    geom_col( data=tmp, aes(x=location, y=distrib, fill=l ), linetype='dashed', size=1, colour='#3399FF' ) +
    scale_fill_manual( name=NULL, values=c('#66CCFF') ) +
    ggpattern::geom_col_pattern( aes(x=shipping_acl_awardee, y=in_transit_total, pattern=vax_manufacturer), fill='#66CCFF', pattern_fill='#66CCFF', colour='#3399FF', pattern_spacing=0.02 ) + 
    ggpattern::scale_pattern_discrete( name=NULL, choices = c("stripe", "crosshatch", "circle"),
                                        labels = c('Janssen','Moderna','Pfizer') ) +
    theme(axis.title.x = element_blank()) +
    ylab('Number of doses in transit') +
    scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")) +
    theme(plot.title = element_text(hjust = 0.5) )

# ggplot(dp) +
#     facet_wrap(~ shipping_acl_awardee, nrow=2, scale='free_y' ) +
#     ggpattern::geom_col_pattern( aes(x=shipping_awardee_category, y=in_transit_total, pattern=vax_manufacturer), fill='#66CCFF', pattern_fill='#66CCFF', colour='#3399FF', pattern_spacing=0.02 ) + 
#     ggpattern::scale_pattern_discrete( name=NULL, choices = c("stripe", "crosshatch", "circle"),
#                                        labels = c('Janssen','Moderna','Pfizer') ) +
#     theme(axis.title.x = element_blank()) +
#     ylab('Number of doses in transit') +
#     scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")) +
#     # ggtitle( paste('Total Doses in Transit on', ref_date)) + 
#     theme(plot.title = element_text(hjust = 0.5) )
```
```{r}
dp <- vxord %>% group_by( shipping_acl_awardee ) %>%
  summarise( in_transit_total = sum( doses_approved[order_approved_date<=latest_history_date & (is.na(first_delivery_date) | first_delivery_date>latest_history_date)] ), .groups='drop_last' ) %>%
  mutate( l = 'total in transit' )

tmp <- stock7d %>% mutate( l = 'addl forecasted distribution' )
ggplot(dp) +
    geom_col( data=tmp, aes(x=location, y=distrib, fill=l ), linetype='dashed', size=0, colour='#3399FF' ) +
    scale_fill_manual( name=NULL, values=c('#66CCFF') ) +
    ggpattern::geom_col_pattern( aes(x=shipping_acl_awardee, y=in_transit_total, pattern=l), fill='#66CCFF', pattern_fill='#66CCFF', size=0, pattern_colour='#3399FF', pattern_spacing=0.02 ) + 
    ggpattern::scale_pattern_manual( name=NULL, values=c('stripe') ) +
    theme(axis.title.x = element_blank()) +
    ylab('Number of doses in transit') +
    scale_y_continuous(label = scales::unit_format(unit = "K", scale = 1e-3, sep = "")) +
    theme(plot.title = element_text(hjust = 0.5) )
```
```{r}
dp <- vxord %>% group_by( shipping_awardee_category, shipping_acl_awardee ) %>%
  summarise( in_transit_total = sum( doses_approved[order_approved_date<=ref_date & (is.na(first_delivery_date) | first_delivery_date>ref_date)] ), .groups='drop_last' )
dp
```
Total shipments per shipping awardee
```{r echo=FALSE}
#str(orders )
print('list of awardees')
unique( orders$awardee )
```

